Investigating soundscapes perception through acoustic scenes simulation.

### 2018-01-19 lafaySimulatedScenePerception.tex
plus
incorporating jfp comments

We propose in this paper to consider the use of a soundscape simulator that the subject can employ to objectify his/her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two abovementioned approaches. As the subject is asked to produce audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the subject's mental representation that is characterized both semantically (the nature of sound sources) and physically (\eg~the levels of sound sources).

@@

In order to account for the possiblity that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only. These two types of sound classes also have specific scheduling procedures during the simulation process.

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

All the subjects perform the experiment on standard desktop computers with the same hardware and software configurations. The audio files are played in diotic conditions using headphones. During the tutorial, subjects are asked to adjust the sound volume to a comfortable level. Once set, they are not allowed to modify it during the remaining of the experiment.

@@

All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.

@@

Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues, and to answer questions.

\subsubsection*{subjects}

@@

For each participant, sound scenes are played in a quasi random order. 5 ideal scenes and 5 non ideal scenes are first sequenced to allow the subjects to calibrate their scores. These first 10 scenes are played back again at the end of the experiment. Only the last evaluations are taken into account.

@@

All the subjects perform the experiment on computers with equivalent hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained by the simulation experiment of experiment 1.a. The output sound level is the same for all the subjects.

@@

\subsubsection*{subjects}

@@

10 students (8 male, 2 female; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment. None of them took part to the previous simulation experiment (experiment 1.a).

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.

\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis. For instance, for the abstraction level 1, that includes 44 classes of sounds ($n = 44$).

@@

@@

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary i-scenes \emph{vs.} ni-scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for i- and ni-scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.

@@

  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
                             \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
\end{figure}

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni). Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

@@

\begin{figure}[t]
  \myfloatalign
  \includegraphics[width=\linewidth]{gfx/pa5_1_en}
  \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level. The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
\end{figure}

@@

Moreover, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$). This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes. Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.

@@

On contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the ni-scenes, shows that this is indeed the absolute level that matters.

@@

%\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir. Les résultats pourraient tout aussi bien s'expliquer de manière plus simple, sans parler de modes de perception différents, en supposant que le volume sonore a toujours un impact négatif sur l'agrément, mais que si un accroissement du volume sonore est concentré sur des marqueurs positifs, l'impact positif de ceux-ci contrebalance l'impact négatif du volume.}

Another hypothesis would be that the volume somehow act as an hedonic "gain" factor. If the volume of a negative marker is high, it lowers the overall pleasantness. If the volume of a positive marker is high, it highers the overall pleasantness. Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform an variance analysis. We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni) and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate whether there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.

@@

\begin{figure}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_13}\label{fig:xp2_Asubject}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene$\mathcal{A}_{scene}$ (b) versus scene type. Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
\end{figure}

@@

@@

This strongly confirms that i-markers do have a positive impact on the perception of an environment. The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}. Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.

@@

\begin{itemize}
\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture;
\item simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
\end{itemize}

>>>>

We propose in this paper to consider the use of a soundscape simulator that the subject can employ to objectify his/her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above mentioned approaches. As the subject is asked to produce audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the subject's mental representation that is characterized both semantically (the nature of sound sources) and physically (\eg~the levels of sound sources).

@@

In order to account for the possibility that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only. These two types of sound classes also have specific scheduling procedures during the simulation process.

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four \ml{Three ?} levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

All the subjects performed the experiment on standard desktop computers with the same hardware and software configurations. The audio files were played in diotic conditions using headphones. During the tutorial, subjects were asked to adjust the sound volume to a comfortable level. Once set, they were not allowed to modify it during the remaining of the experiment.

@@

All the subjects performed the experiment at the same time. They were equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.

@@

Three experimenters (one in each room) were available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues, and to answer questions.

\subsubsection*{Subjects}

@@

For each participant, sound scenes are played in a quasi random order. 5 ideal scenes and 5 non ideal scenes are first sequenced to allow the subjects to calibrate their scores. These first 10 scenes are played back again at the end of the experiment. Only the last evaluations are taken into account. Each participant evaluates all the sound scenes.

@@

All the subjects perform the experiment on computers with equivalent hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained in experiment 1.a. The output sound level is the same for all the subjects.

@@

\subsubsection*{Subjects}

@@

10 students (8 male, 2 female; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} took part to the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment. None of them took part to the previous simulation experiment (experiment 1.a).

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores. $\mathcal{A}_{subject}$ is computed for i-scenes and ni-scenes.

\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis. For instance, for the abstraction level 1, that includes 44 classes of sounds, the dimension is thus $44$ ($n = 44$).

@@

\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène
a quoi sert Asubject ici ?}

@@

\jfp{a la limite on pourrait ne pas écarter arbitrairement de scène et ne garder que celles pour lesquelles les iscenes sont signif. + plaisante que les ni-scene. ca revient au même. c'est les test qui les écarte. il faudrait peut être justifier pourquoi on fait cela  : pour ne pas fausser les résultats de l'analyse et vérifier que les i scènes sont effectivement perçues + plaisantes que les ni.}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary i-scenes \emph{vs.} ni-scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for i- and ni-scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for i-scenes.

@@

  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
                             \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$. \jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?}}\label{fig:soundsource}
\end{figure}

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni). For the 36 i and ni-scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value \ml{ shown between parenthesis}. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

@@

\begin{figure}[t]
  \myfloatalign
  \includegraphics[width=\linewidth]{gfx/pa5_1_en}
  \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level. The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$. Baseline results are achieved by considering random vectors as input.}\label{fig:pa5}
\end{figure}

@@

Moreover, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$). This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes. Sound level of textures seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.

@@

On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the ni-scenes, shows that this is indeed the absolute level that matters.

@@

\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir. Les résultats pourraient tout aussi bien s'expliquer de manière plus simple, sans parler de modes de perception différents, en supposant que le volume sonore a toujours un impact négatif sur l'agrément, mais que si un accroissement du volume sonore est concentré sur des marqueurs positifs, l'impact positif de ceux-ci contrebalance l'impact négatif du volume.}

\ml{Ma traduction : Another hypothesis would be that the volume somehow act as an hedonic "gain" factor. If the volume of a negative marker is high, it lowers the overall pleasantness. If the volume of a positive marker is high, it highers the overall pleasantness. Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.}

@@

The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness. To do so, we perform a variance analysis. We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni) and the presence/absence of markers (km/rm). As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate whether there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analyses are done using the Tukey-Kramer procedure. \jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.}

@@

\begin{figure}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_13}\label{fig:xp2_Asubject}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene$\mathcal{A}_{scene}$ (b) versus scene type. Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7. \ml{maj figure: sujet->subject, et changer am et sm pour km et rm}}\label{fig:xp2A}
\end{figure}

@@

\jfp{je crois que l'analyse post-hoc n'est pas nécessaire lorsqu'il n'y a que 2 niveaux pour les facteurs de l'ANOVA. les conclusions seront strictement les mêmes que pour l'ANOVA.c'est utile à partir de 3 niveaux, pour savoir quelle moyenne diffère signif. des autres.}

@@

This strongly confirms that i-markers do have a positive impact on the perception of an environment. The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}. Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.

@@

\begin{itemize}
\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture;
\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
\end{itemize}
